---
title: "Comparison for W3 and 13"
author: "Félix Báez-Santiago, Valeria Castañeda-Saucedo, Xinyu Lin and Angelina Lu"
date: "2025-11-13"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(ggplot2)
library(clue)
```

```{r}
m_w3_nocov  <- readRDS("lca_w3_k4_nocov.rds")
m_w13_nocov <- readRDS("lca_w13_k4_nocov_update.rds")
m_w3_cov    <- readRDS("lca_w3_k4_cov.rds")
m_w13_cov   <- readRDS("lca_w13_k4_cov_update.rds")
```

```{r}
# Fit comparison (AIC/BIC/G^2, N, K)
fit_row <- function(m, wave, cov){
  tibble(
    wave     = wave,
    covars   = cov,
    k        = ncol(m$P),  
    N        = m$N,
    AIC      = m$aic,
    BIC      = m$bic,
    G2       = m$Gsq
  )
}
fit_cmp <- bind_rows(
  fit_row(m_w3_nocov,  "W3",  "no"),
  fit_row(m_w13_nocov, "W13", "no"),
  fit_row(m_w3_cov,    "W3",  "yes"),
  fit_row(m_w13_cov,   "W13", "yes")
)
fit_cmp
```

```{r}
# Class size comparison (estimated mixing proportions)
mix_row <- function(m, wave, cov){
  tibble(wave = wave, covars = cov, class = seq_along(m$P), proportion = as.numeric(m$P))
}
mix_cmp <- bind_rows(
  mix_row(m_w3_nocov,  "W3",  "no"),
  mix_row(m_w13_nocov, "W13", "no"),
  mix_row(m_w3_cov,    "W3",  "yes"),
  mix_row(m_w13_cov,   "W13", "yes")
)
mix_cmp

```

```{r}
# Quick entropy / classification sharpness
entropy <- function(p) -sum(p * log(p + 1e-12))
norm_entropy <- function(post) {
  # average normalized entropy across cases (0 = sharp, 1 = diffuse)
  K <- ncol(post); Hmax <- log(K)
  mean(apply(post, 1, entropy) / Hmax)
}
sharpness <- function(post) mean(apply(post, 1, max))  # closer to 1 is better

qual_cmp <- tibble(
  wave   = c("W3","W13","W3","W13"),
  covars = c("no","no","yes","yes"),
  norm_entropy = c(norm_entropy(m_w3_nocov$posterior),
                   norm_entropy(m_w13_nocov$posterior),
                   norm_entropy(m_w3_cov$posterior),
                   norm_entropy(m_w13_cov$posterior)),
  max_p_mean   = c(sharpness(m_w3_nocov$posterior),
                   sharpness(m_w13_nocov$posterior),
                   sharpness(m_w3_cov$posterior),
                   sharpness(m_w13_cov$posterior))
)
qual_cmp
```

We evaluated classification quality using normalized entropy (lower = better) and the mean maximum posterior probability (higher = better).

For Wave 3, the no-covariate model showed clearer class separation than the covariate model (entropy: 0.215 vs 0.308; max posterior: 0.891 vs 0.851). This suggests that covariates introduced additional variability and slightly diluted classification certainty. In contrast, for Wave 13, adding covariates improved classification quality (entropy: 0.245 to 0.207; max posterior: 0.867 to 0.886), indicating that class membership in Wave 13 is more strongly structured by the covariates.

**Do the proportions of respondents in each class differ between 1996 and 2016?**

Without covariances

```{r}
match_classes <- function(m_w3, m_w13){
  
  flatten_probs <- function(m){
    mats <- lapply(m$probs, function(x) as.matrix(x))
    P <- do.call(cbind, mats)
    rownames(P) <- paste0("c", seq_len(nrow(P)))
    P
  }
  
  # get flattened matrices
  P_w3  <- flatten_probs(m_w3)
  P_w13 <- flatten_probs(m_w13)
  
  # cosine similarity
  cos_sim <- function(A, B){
    A <- as.matrix(A); B <- as.matrix(B)
    num <- A %*% t(B)
    A_n <- sqrt(rowSums(A^2)); B_n <- sqrt(rowSums(B^2))
    sweep(sweep(num, 1, A_n, "/"), 2, B_n, "/")
  }
  
  S <- cos_sim(P_w3, P_w13)
  S[is.na(S)] <- 0
  
  # Hungarian: convert similarity to cost
  cost <- max(S) - S
  assign <- clue::solve_LSAP(cost)
  
  matched_pairs <- tibble(
    w3_class   = seq_len(nrow(S)),
    w13_class  = as.integer(assign),
    similarity = S[cbind(seq_len(nrow(S)), as.integer(assign))]
  )
  
  return(matched_pairs)
}
```

```{r}
compare_props <- function(matched_pairs, m_w3, m_w13){
  p_w3 <- tibble(
    w3_class = seq_along(m_w3$P),
    prop_w3  = as.numeric(m_w3$P)
  )
  
  p_w13 <- tibble(
    w13_class = seq_along(m_w13$P),
    prop_w13  = as.numeric(m_w13$P)
  )
  
  matched_pairs %>%
    left_join(p_w3,  by = "w3_class") %>%
    left_join(p_w13, by = "w13_class")
}

```

```{r}
# no covariates
matched_nocov <- match_classes(m_w3_nocov, m_w13_nocov)
prop_nocov    <- compare_props(matched_nocov, m_w3_nocov, m_w13_nocov)

prop_nocov
```

To compare latent classes across waves, we aligned Wave 3 and Wave 13 classes based on cosine similarity of their item-response probability profiles.\

High similarity indicates that the classes represent very similar underlying subpopulations.

With covariance

```{r}
# with covariates
matched_cov <- match_classes(m_w3_cov, m_w13_cov)
prop_cov    <- compare_props(matched_cov, m_w3_cov, m_w13_cov)

prop_cov
```

**Which demographic covariates predict class membership, and do these associations differ across years?**

```{r}
m_w3_cov$coeff
m_w13_cov$coeff
```

```{r}
m_w3_cov$probs
m_w13_cov$probs
```

```{r}
m_w3_nocov$probs
m_w13_nocov$probs
```

**How can we assess the model fit using normalized entropy and the average maximum posterior probability?**

```{r}

entropy_row <- function(p_row){
  -sum(p_row * log(p_row + 1e-12))
}

norm_entropy_model <- function(post_mat){
  K <- ncol(post_mat)
  Hmax <- log(K)
  H_i <- apply(post_mat, 1, entropy_row) 
  mean(H_i / Hmax)        
}

sharpness_model <- function(post_mat){
  mean(apply(post_mat, 1, max))
}

class_quality <- function(mod, wave, covars){
  post <- mod$posterior
  tibble(
    wave         = wave,
    covars       = covars,   # "yes" / "no"
    n            = nrow(post),
    K            = ncol(post),
    norm_entropy = norm_entropy_model(post),
    max_p_mean   = sharpness_model(post)
  )
}

```

```{r}
qual_cmp <- bind_rows(
  class_quality(m_w3_nocov,  wave = "W3",  covars = "no"),
  class_quality(m_w13_nocov, wave = "W13", covars = "no"),
  class_quality(m_w3_cov,    wave = "W3",  covars = "yes"),
  class_quality(m_w13_cov,   wave = "W13", covars = "yes")
)

qual_cmp

```

```{r}
library(ggplot2)

ggplot(qual_cmp, aes(x = wave, y = norm_entropy, 
                     group = covars, linetype = covars)) +
  geom_line() +
  geom_point() +
  labs(x = "Wave",
       y = "Normalized entropy",
       linetype = "Covariates",
       title = "Classification quality (entropy) across waves and models")

```

```{r}

```
